Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-06-18 paper/main.tex
done with Section 3

Notice, in the center/right panels of Figure \ref{fig:plot_data}, the precision of each
low/high resolution spectra increases as $k$ increases.  The curves get progressively 
tighter from left to right.  We would like to conveniently wrap this domain-specific 
knowledge inside our Bayesian framework later.  To this end, \cite{moran2023mira} 
used the multiple low- and high-resolution spectra across all cosmologies to
obtain precision estimates, $p_1,\dots, p_n$, across the $n$ 
wavenumber values using a log-log regression model. A multiplier $c\approx 3.73$ for 
the increase in precision from the low- to high- resolution output is also 
estimated with this regression model. 
We use these estimates to define three 
diagonal precision matrices, one for each data type (perturbation theory, 
low resolution average, and high resolution). Specifically, denote $\Lambda_p$, 
$\Lambda_\ell$, and $\Lambda_h$ as the $n\times n$ diagonal matrices with elements
\[
\Lambda_p^{(ii)} = \begin{cases}
    10^5 &\text{for } k_i < 0.04 \\
    0  &\text{otherwise}\\
    \end{cases}
\quad
\Lambda_\ell^{(ii)} = \begin{cases}
    16p_i &\text{for } 0.04 \leq k_i < 0.25 \\
    0  &\text{otherwise}\\
    \end{cases}
\quad
\Lambda_h^{(ii)} = \begin{cases}
    cp_i &\text{for } 0.04 \leq k_i < 5 \\
    0  &\text{otherwise}\\
    \end{cases}
\]
for $i=1,\dots, n$.  In $\Lambda_p$, we use the sufficiently high precision 
of $10^5$ to impose the deterministic nature of the perturbation theory output.
In $\Lambda_\ell$, the multiplication by 16 accounts for the sample size of the
low resolution runs.
We then calculate a weighted average for each cosmology: 
$\bar y = \Lambda^{-1}(\Lambda_p y_p + \Lambda_{\ell} \bar{y}_\ell + \Lambda_h y_h)$, 
where $\Lambda = \Lambda_p + \Lambda_\ell + \Lambda_h$ and 
$\bar{y}_\ell = \frac{1}{16}\sum_{r=1}^{16} y_{\ell_r}$. In Figure \ref{fig:plot_data},
the weighted average is shown in dashed orange.  As an estimate of the true
underlying spectrum, the weighted average fluctuates too drastically.  
This is especially apparent after centering (right panel).
\textcolor{red}{(@Steve - how are we centering?)}
Although the weighted average incorporates all available observatons and 
expert knowledge regarding the precisions, it fails to account for the spatial 
dependence inherent in the smooth but stochastic spectra, which is pivotal to 
effectively inferring the true spectrum.

@@

From plots of the multiple low-resolution spectra (e.g., Figure \ref{fig:plot_data}), 
we find they vary smoothly about some mean $S$, suggesting that a dense covariance 
matrix $\Sigma_\varepsilon$ (accounting for domain knowledge and spatial dependence) 
may outperform a diagonal covariance $\Lambda_\ell^{-1}$ for $\bar Y$. From Section 
\ref{sec:data}, we can incorporate domain knowledge from $\Lambda_p, \Lambda_\ell,$ 
and $\Lambda_h$ within our model, and calculate the weighted average for each cosmology: 
$\bar y = \Lambda^{-1}(\Lambda_p y_p + \Lambda_{\ell} \bar{y}_\ell + \Lambda_h y_h)$, 
where $\Lambda = \Lambda_p + \Lambda_\ell + \Lambda_h$. 
In previous work, we have modeled multiple candidates for 
$\Sigma_\varepsilon$ \citep{walsh2023bayesian} and found that a Mat\'ern covariance 
$\Sigma_\ell$ trained on the low-resolution runs (pre-scaled by the precision 
values and subtracting a LOESS-smoothed average) performs the best. 
We specify this structure within $\Sigma_\varepsilon$ and refer readers 
to \cite{walsh2023bayesian} for a more detailed exposition. This yields 
$\Sigma_\varepsilon=\left(\Lambda_p^{-1} + \Sigma_\ell + \Lambda_h^{-1}\right)^{-1}$, 
where $\Lambda_p, \Sigma_\ell^{-1}, \text{ and } \Lambda_h$ are all $n\times n$ 
and are 0 wherever the corresponding data type is biased. From here, we estimate 
$\tau_\ell, \theta_\ell$ with maximum likelihood, and input these within 
$\Sigma_\varepsilon$, which we use as an input to our model.

To train our hierarchical model, we initialize our hyperparameter estimates across 
all cosmologies by modeling the first (M001) with 50,000 MCMC runs, and use these 
as initial values for all cosmologies. Specifically, we save the last MCMC 
iteration of $W$, $\theta_w$ and $\theta_S$ as $W_0$, $\theta_{w0}$ and $\theta_{S0}$ 
respectively and use each as starting values for all 111 training cosmology fits, 
where each cosmology is fit with 10,000 MCMC runs (5,000 removed as burn-in and 
every 5th sample is kept, resulting in 1,000 posterior samples).

From these samples, we can estimate and quantify the uncertainty within each 
cosmology's average and latent layer $W$. Figure \ref{fig:plot_fit} shows an 
example of the uncertainty for our posterior mean for the first (of 111) training 
cosmologies; the posterior mean is subtracted in order to visualize this uncertainty. 
The UQ from the DGP.FCO model accounts for varying precisions across differing $k$ 
values, as well as the spatial correlation within the functional realizations.
Figure \ref{fig:plot_warp} provides elliptical slice samples of $W$ along with the 
average warping. From these samples, we see some departure from stationarity 
estimated in the higher $X$ (i.e., scaled $k$) values.

\begin{figure}[ht]
    \centering
    \includegraphics[width=6in]{plot_fit.png}
    \caption{95\% credible intervals for the power spectrum of model 1 (posterior mean 
             subtracted). Plots of the perturbation theory, low-resolution and high-resolution 
             runs are also shown, alongside the weighted average.}
    \label{fig:plot_fit}
\end{figure}

\begin{figure}[ht]
   \centering
   \includegraphics[width=6in]{plot_warp_M001.png}
   \caption{Left: MCMC samples of $W$ for model M001 after burnin and thinning. 
            Right: Corresponding 95\% credible intervals for $W$, where departure 
            from the identity line represents nonstationary behavior.
            \textcolor{red}{(This plot needs to be made fresh (not using the package
             function.  Adjust y-axis.  We don't need the yellow/red coloring.)}}
   \label{fig:plot_warp}
\end{figure}

>>>>

Notice, in the center/right panels of Figure \ref{fig:plot_data}, the precision of each
low/high resolution spectra increases as $k$ increases.  The curves get progressively 
tighter from left to right.  We would like to conveniently wrap this domain-specific 
knowledge inside our Bayesian framework later.  To this end, \cite{moran2023mira} 
used the multiple low- and high-resolution spectra across all cosmologies to
obtain precision estimates, $p_1,\dots, p_n$, across the $n$ 
wavenumber values using a log-log regression model. A multiplier $c\approx 3.73$ for 
the increase in precision from the low- to high- resolution output is also 
estimated with this regression model. 
We use these estimates to define three 
diagonal precision matrices, one for each data type (perturbation theory, 
low resolution average, and high resolution). Specifically, denote $\Lambda_p$, 
$\Lambda_\ell$, and $\Lambda_h$ as the $n\times n$ diagonal matrices with elements
\begin{equation}\label{eq:lambda}
\Lambda_p^{(ii)} = \begin{cases}
    10^5 &\text{for } k_i < 0.04 \\
    0  &\text{otherwise}\\
    \end{cases}
\quad
\Lambda_\ell^{(ii)} = \begin{cases}
    16p_i &\text{for } 0.04 \leq k_i < 0.25 \\
    0  &\text{otherwise}\\
    \end{cases}
\quad
\Lambda_h^{(ii)} = \begin{cases}
    cp_i &\text{for } 0.04 \leq k_i < 5 \\
    0  &\text{otherwise}\\
    \end{cases}
\end{equation}
for $i=1,\dots, n$.  In $\Lambda_p$, we use the sufficiently high precision 
of $10^5$ to impose the deterministic nature of the perturbation theory output.
In $\Lambda_\ell$, the multiplication by 16 accounts for the sample size of the
low resolution runs.
We then calculate a weighted average for each cosmology: 
$\bar y = \Lambda^{-1}(\Lambda_p y_p + \Lambda_{\ell} \bar{y}_\ell + \Lambda_h y_h)$, 
where $\Lambda = \Lambda_p + \Lambda_\ell + \Lambda_h$ and 
$\bar{y}_\ell = \frac{1}{16}\sum_{r=1}^{16} y_{\ell_r}$. In Figure \ref{fig:plot_data},
the weighted average is shown in dashed orange.  As an estimate of the true
underlying spectrum, the weighted average fluctuates too drastically.  
This is especially apparent after centering (right panel).
\textcolor{red}{(@Steve - how are we centering?)}
Although the weighted average incorporates all available observatons and 
expert knowledge regarding the precisions, it fails to account for the spatial 
dependence inherent in the smooth but stochastic spectra, which is pivotal to 
effectively inferring the true spectrum.

@@

Here, we revisit the Mira-Titan suite described in Section \ref{sec:data} to provide
final details for our estimation of the true (but unknown) matter power spectrum for
each cosmology.  Recall, with $\Lambda_p$, $\Lambda_\ell$, and $\Lambda_h$ as 
defined in Eq.~(\ref{eq:lambda}), we compute
$\bar y = \Lambda^{-1}(\Lambda_p y_p + \Lambda_{\ell} \bar{y}_\ell + \Lambda_h y_h)$, 
where $\Lambda = \Lambda_p + \Lambda_\ell + \Lambda_h$.
Turning to $\Sigma_\epsilon$, since the low-resolution spectra 
vary smoothly about $S$ (e.g., Figure \ref{fig:plot_data}), we prefer a dense 
covariance ($\Sigma_\ell$) for wavelengths $0.04 \leq k < 0.25$.
In previous work, we considered multiple candidates for 
$\Sigma_\varepsilon$ \citep{walsh2023bayesian} and found that a Mat\'ern covariance 
for $\Sigma_\ell$ trained on the low-resolution runs (pre-scaled by the precision 
values and subtracting a LOESS-smoothed average) performs the best. 
We specify this structure within $\Sigma_\varepsilon$ and refer readers 
to \cite{walsh2023bayesian} for a more detailed exposition. This yields 
$\Sigma_\varepsilon=\left(\Lambda_p^{-1} + \Sigma_\ell + \Lambda_h^{-1}\right)^{-1}$.

We then fit our DGP.FCO model on each of 111 training cosmologies, reserving 6 cosmologies
for later hold-out testing.  To speed-up burn-in of MCMC chains, we conduct
50,000 MCMC iterations for the first cosmology, then initialize the chains for all 
other cosmologies with these burned-in samples, so fewer iterations are needed for the
remaining 110 cosmologies.  Ultimately, we conduct 10,000 iterations for each of these,
removing 5,000 for burn-in, then thinning by 5.  Figure \ref{fig:plot_fit} 
visualizes the DGP.FCO fit for the first training cosmology.  We center the plot
by subtracting the posterior mean in order to more easily visualize the credible
interval bounds.  The UQ from the DGP.FCO model accounts for varying precisions 
across differing $k$ values, as well as the spatial correlation within the 
functional realizations.  To provide insight into the nonstationarity of the power
matter spectrum, Figure \ref{fig:plot_warp} shows the elliptical slice samples of $W$.
These warpings depart from the identiy mapping around $x=0.4$ \textcolor{red}{(@Steve - 
please change the x-axis to match the actual x-axis values, then adjust this number)}, 
stretching inputs where dynamics are shifting more rapidly for larger $k$ values.

\begin{figure}[ht]
    \centering
    \includegraphics[width=6in]{plot_fit.png}
    \caption{95\% credible intervals for the power spectrum of the first Mira-Titan
    cosmology (centered by the predicted posterior mean). Perturbation theory, 
    low-resolution, and high-resolution runs are shown for context.
    \textcolor{red}{(@Steve - combine this plot with Figure 6.  The two panels of Figure 6
    are redundant and unecessary (yet oddly, they don't match?).  Have this Mira-Titan 
    plot take up 2/3 of the space, then have just the right panel from Figure 6 take 
    up the remaining 1/3.  Update the figure caption and referencing accordingly.)}}
    \label{fig:plot_fit}
\end{figure}

\begin{figure}[ht]
   \centering
   \includegraphics[width=6in]{plot_warp_M001.png}
   \caption{Left: MCMC samples of $W$ for model M001 after burnin and thinning. 
            Right: Corresponding 95\% credible intervals for $W$, where departure 
            from the identity line represents nonstationary behavior.
            \textcolor{red}{(remove this figure, combine right panel with previous)}}
   \label{fig:plot_warp}
\end{figure}
